---
title: "Beer!"
author: "Jasmine Dumas"
date: "October 28, 2015"
output: word_document
---
1. **Data source**:
```{r}
beer = read.table("http://www.craftbeeranalytics.com/uploads/3/3/8/9/3389428/ratebeer_beerjobber.txt", header=TRUE, sep="")
head(beer)
colnames(beer) # we'd be predicting column score.overall using abv, ratings, & score.by.style
```

2. **Data Clean**:
```{r}
beer <- na.omit(beer) 
```

3. **Research Question**: What contributes to beer rating scores overall, abv, style, brewer?   
4. **Dependent Variable**: score.overall   
5. **Independent Variable**: score.by.style, name, brewer, style, abv, ratings    
6. **Quantitative vs. Qualitative**: The quantitative variables are score.overall, abv, ratings,score.by.style. The qualitative variable is name and brewer.  
7. **Correlation and Scatter plots**: 
```{r}
# no multicollinearity between variables - want to be low!
cor(beer$abv, beer$ratings)
cor(beer$abv, beer$score.by.style)
cor(beer$ratings, beer$score.by.style)

# want correlation between response and covariates - want to be high!
cor(beer$score.overall, beer$abv)
cor(beer$score.overall, beer$ratings)
cor(beer$score.overall, beer$score.by.style)

# what do these variables all look like against the response y variable
library(ggplot2)
ggplot(beer, aes(y = score.overall, x = abv)) + geom_point() # log?
ggplot(beer, aes(y = score.overall, x = ratings)) + geom_point() # log?
ggplot(beer, aes(y = score.overall, x = score.by.style)) + geom_point() # weird-linear?
# these graphs look weird maybe try inverse or logrithm scale and re-graph?
```

8. **Building the model**:
```{r}
beer.model1 = lm(score.overall ~ abv + ratings + score.by.style + brewer, data=beer) # low F-statistic
summary(beer.model1)
# re-run model
beer.model2 = lm(score.overall ~ abv + ratings + score.by.style, data=beer) # better significant results
summary(beer.model2)
# another run with style as a dummy variable
beer.model3 = lm(score.overall ~ abv + ratings + score.by.style + style, data=beer)
summary(beer.model3)
# stepwise regression for the final beer.model
full.beer = step(beer.model3, direction="both")
```



